
---

title: "Assignment 3"
author: "Jakku Keerthi Priya"
date: "2026-03-03"
output:
  pdf_document:
    latex_engine: xelatex
---

Necessary Packages
```{r}
library(caret)
library(e1071)
set.seed(123)
```


Load Data
```{r}
UniversalBank<-read.csv("~/Desktop/R documents/Assignment files/FML_Assignment2/UniversalBank.csv")
```


```{r}
head(UniversalBank)
```


Summary of Data

```{r}
summary(UniversalBank)
```


Partitioning the data into training (60%) and validaƟon (40%) sets

```{r}
indextrain<- createDataPartition(UniversalBank$Personal.Loan, p=0.6, list=FALSE)

training<- UniversalBank[indextrain, ]

validation<- UniversalBank[-indextrain, ]

```



Create a pivot table

```{r}
  pivot_table <- table(CreditCard=training$CreditCard, PersonalLoan=training$Personal.Loan,Online=training$Online)
  
  pivot_table
```


Calculating probability

```{r}
loan_accepted_online_CC <- pivot_table["1","1","1"]
total_CC_online <- sum(pivot_table["1", ,"1" ])
probability_loanaccepted <- loan_accepted_online_CC/total_CC_online
probability_loanaccepted
```




Loan (rows) as a function of Online (columns) : Pivot table

```{r}
loan_online_pivot  <- table(PersonalLoan = training$Personal.Loan,
                            Online = training$Online)
loan_online_pivot
```


Loan (rows) as a function of Credit card (columns) : Pivot table

```{r}
loan_CC_pivot<- table(PersonalLoan = training$Personal.Loan,
                        CreditCard = training$CreditCard)
loan_CC_pivot
```


Compute the following quantities



1. P(CC = 1 | Loan = 1) (the proportion of credit card holders among the loan acceptors) 

```{r}
CCholders_loan_accepted <- loan_CC_pivot["1","1"] / sum(loan_CC_pivot["1", ])
CCholders_loan_accepted
```



2. P(Online = 1 | Loan = 1) 

```{r}
onlineuser_loan_accepted <-  loan_online_pivot["1","1"] / sum(loan_online_pivot["1", ])

onlineuser_loan_accepted
```



3. P(Loan = 1) (the proportion of loan acceptors) 

```{r}
loan_accepted <- sum(training$Personal.Loan == 1) / nrow(training)
loan_accepted

```



4. P(CC = 1 | Loan = 0) 

```{r}
Cardholder_loanrejected <- loan_CC_pivot["0","1"] / sum(loan_CC_pivot["0", ])
Cardholder_loanrejected

```




5. P(Online = 1 | Loan = 0) 

```{r}
inactiveuser_loanrejected <- loan_online_pivot["0","1"] / sum(loan_online_pivot["0", ])

inactiveuser_loanrejected
```




6. P(Loan = 0) 

```{r}
loan_rejected <- sum(training$Personal.Loan == 0) / nrow(training)

loan_rejected
```




Naive Bayes probability P(Loan = 1 | CC = 1, Online = 1 )

```{r}
CCholder_onlineuser <- nrow(training[training$CreditCard == 1 & training$Online  == 1, ]) / nrow(training)

Loanaccepted_CCholder_onlineuser <- (CCholders_loan_accepted*onlineuser_loan_accepted*loan_accepted)/CCholder_onlineuser

Loanaccepted_CCholder_onlineuser
                            
                            
                            
                            
```




Comparing the pivot table value with the Naive Bayes Probability

```{r}
probability_loanaccepted 
```

```{r}
Loanaccepted_CCholder_onlineuser
```




There is a slight differences between the two values. The Pivot table value looks more accurate as it is without any assumptions and Naive Bayes assumes independence between predictors.





Naive Bayes model

```{r}
model <- naiveBayes(Personal.Loan ~ CreditCard + Online, data = training)

predict_probability <- predict(model, training , type = "raw")
NaiveBayes_probability <- predict_probability[training$CreditCard == 1 & training$Online == 1, 2]
NaiveBayes_probability

```





Analysis:
Naive Bayes probability is more accurate than the probability form Section E
because model-derived probability considers distribution and relationship within the data.


















