---
title: "Assignment_2"
author: "Keerthi Priya Jakku"
date: "2026-02-18"
output: 
  pdf_document:
    latex_engine: lualatex
---

Necessary packages
```{r}

library(ggplot2)
library(dplyr)
library(lattice)
library (caret)
library(class)

set.seed(123)

```


Load Data
```{r}
UniversalBank<-read.csv("~/Desktop/R documents/Assignment files/FML_Assignment2/UniversalBank.csv")
```


Converting Education to Dummy variable

```{r}
UniversalBank$Education <- as.factor(UniversalBank$Education)

dummy_model <- dummyVars(~ Education, data = UniversalBank)
dummy_data <- predict(dummy_model, newdata = UniversalBank)

UniversalBank <- UniversalBank[, !colnames(UniversalBank) %in% "Education"]
UniversalBank <- cbind(UniversalBank, dummy_data)

```


Excluding ID and Zip Code
```{r}
UniversalBank<- UniversalBank[, !(names(UniversalBank) %in% c("ID", "ZIP.Code"))]
```

Converting Personal Loan to factor
```{r}
UniversalBank$Personal.Loan <- as.factor(UniversalBank$Personal.Loan)

```


Renaming the column names
```{r}
colnames(UniversalBank)[colnames(UniversalBank) == "EducationEducation_1"] <- "Education_1"
colnames(UniversalBank)[colnames(UniversalBank) == "EducationEducation_2"]<- "Education_2"
colnames(UniversalBank)[colnames(UniversalBank) == "EducationEducation_3"]<- "Education_3"
```



Partitioning the data into training (60%) and validaƟon (40%) sets
```{r}
indextrain<-createDataPartition(UniversalBank$Personal.Loan,
                                p=0.6,
                                list=FALSE)
training<-UniversalBank[indextrain, ]
validation<- UniversalBank[-indextrain, ]
```


Seperating predictors and target
```{r}
trainingX <- training[, !colnames(training) %in% "Personal.Loan"]
trainingY <- training$Personal.Loan

validationX <- validation[, !colnames(validation) %in% "Personal.Loan"]
validationY <- validation$Personal.Loan
```




Convert predictors to numeric
```{r}

trainingX <- data.frame(lapply(trainingX, as.numeric))
validationX <- data.frame(lapply(validationX, as.numeric))

```


Standardizing the data for k-NN classification
```{r}
prepro <- preProcess(trainingX, method = c("center", "scale"))

trainingX_scaled <- predict(prepro, trainingX)
validationX_scaled <- predict(prepro, validationX)
```



K-NN Classification Using K = 1
```{r}
Predicted_KNN1 <- knn(train = trainingX_scaled,
                      test = validationX_scaled,
                      cl = trainingY,
                      k = 1)

confusionMatrix(Predicted_KNN1, validationY, positive = "1")
```

Finding the best K
```{r}
accuracy <- c()

for(i in 1:20){
  pred <- knn(train = trainingX_scaled,
              test = validationX_scaled,
              cl = trainingY,
              k = i)
  
  accuracy[i] <- mean(pred == validationY)
}

accuracy_table <- data.frame(
  k = 1:10,
  Accuracy = round(accuracy, 6)
)

print(accuracy_table)

accuracy
best_k <- which.max(accuracy)
best_k
```



New customer data

```{r}
new_customer <- data.frame(
  Age = 40,
  Experience = 10,
  Income = 84,
  Family = 2,
  CCAvg = 2,
  Mortgage = 0,
  Securities.Account = 0,
  CD.Account = 0,
  Online = 1,
  CreditCard = 1,
  Education.1 = 0,
  Education.2 = 1,
  Education.3 = 0
)

new_customer_scaled <- predict(prepro, new_customer)

knnpredict <- knn(train = trainingX_scaled,
              test = new_customer_scaled,
              cl= trainingY,
              k = best_k)

knnpredict
```
Outcome: New Customer will not accept the Personal Loan




K-NN Classification Using k = 2
```{r}
Predicted_KNN2 <- knn(train = trainingX_scaled,
                      test = validationX_scaled,
                      cl = trainingY,
                      k = 2)

conf_matrix_k2 <- confusionMatrix(Predicted_KNN2, validationY, positive = "1")
conf_matrix_k2

prediction2 <- knn(trainingX_scaled,
                   new_customer_scaled,
                   cl = trainingY,
                   k = 2)
```


Confusion Matrix Using  k=2
```{r}
knn_k2 <- knn(train = trainingX_scaled,
              test = validationX_scaled,
              cl = trainingY,
              k = 2)

conf_matrix_k2 <- confusionMatrix(knn_k2, validationY, positive = "1")
conf_matrix_k2
```





Dividing the data (50% training / 30% validation / 20% training)

```{r}
indextrain2 <- createDataPartition(UniversalBank$Personal.Loan,
                                    p = 0.5,
                                    list = FALSE)

train2 <- UniversalBank[indextrain2, ]
temp <- UniversalBank[-indextrain2, ]

index_valid2 <- createDataPartition(temp$Personal.Loan,
                                    p = 0.6,
                                    list = FALSE)

valid2 <- temp[index_valid2, ]
test2 <- temp[-index_valid2, ]
```


Seperate predictors and target

```{r}
train2X <- data.frame(lapply(train2[, !colnames(train2) %in% "Personal.Loan"], as.numeric))
train2Y <- train2$Personal.Loan

valid2X <- data.frame(lapply(valid2[, !colnames(valid2) %in% "Personal.Loan"], as.numeric))
valid2Y <- valid2$Personal.Loan

test2X <- data.frame(lapply(test2[, !colnames(test2) %in% "Personal.Loan"], as.numeric))
test2Y <- test2$Personal.Loan

```


Scale predictors
```{r}
prepro2 <- preProcess(train2X, method = c("center", "scale"))
train2X_scaled <- predict(prepro2, train2X)
valid2X_scaled <- predict(prepro2, valid2X)
test2X_scaled <- predict(prepro2, test2X)

```


k-nn predictions
```{r}
train2_pred <- knn(train2X_scaled, train2X_scaled, train2Y, k = best_k)
valid2_pred <- knn(train2X_scaled, valid2X_scaled, train2Y, k = best_k)
test2_pred <- knn(train2X_scaled, test2X_scaled, train2Y, k = best_k)

```


Confusion matrices
```{r}
confusionMatrix(train2_pred, train2Y, positive = "1")
confusionMatrix(valid2_pred, valid2Y, positive = "1")
confusionMatrix(test2_pred, test2Y, positive = "1")
```

Analysis: Thes dataset was split into 60% training and 40% validation sets after converting the Education variable into dummy variables and removing ID and ZIP code. When k = 1 was used, the model achieved a validation accuracy of about 96.45%, and the new customer was predicted as 0, meaning they are unlikely to accept the  loan. After testing different k values from 1 to 20, k = 1 gave the highest accuracy and was chosen as the best value. However, k = 2 was also tested to reduce possible overfitting. Although its accuracy was slightly lower (around 95–96%), the performance was still good and the new customer was again classified as 0. When the data was repartitioned into 50% training, 30% validation, and 20% test sets, the training accuracy was very high while the validation and test accuracies were both around 95%. This suggests that k = 1 may slightly overfit the training data, while k = 2 provides a more stable model. Overall, the k-NN model performs well in predicting loan acceptance, and the customer is predicted not to accept the loan.
 